Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

Defining a language by building an interpreter for it can be traced to the very
early time of programming languages research \cite{landin1966next, reynolds1972definitional}.
Nowadays, even an undergraduate student in computer science is able to build toy languages
through interpreters.
But building a sound abstract interpreter remains as an esoteric and difficult task until recent years.

\citeauthor{van2012systematic} proposed Abstracting Abstract Machines (AAM) methodology which can used
for constructing sound abstract interpreters of higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

Broadly speaking, abstract abstract machines and abstract definitional
interpreters are both some kind of abstract interpreters.
They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interpreters, respectively.
An interesting question would be how we can interderive these abstract semantics artifacts
from the other one.

In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters is intensively studied by Danvy and his
collaborators in the past years
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The abstract machines implement structural operational semantics in continuation style,
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines,
and by defunctionalizing the higher-order functions, one may obtain abstract machines in the
reverse direction.

On the contrary, in terms of their abstract semantics artifacts,
the relations between small-step abstract abstract machines and big-step abstract
definitional interpreters, as well as the question of deriving one from the other is not
well-studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we practically answer the question and relate AAM and ADI
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.

In addition, the abstract abstract machines with unbounded stack naturally correspond
to abstract definitional interpreters. We show after refunctionalizing an AAM with unbounded
stack, and with the proper caching algorithm, the pushdown control-flow analysis can be
obtained.

@@

\subsection{Contribution}

@@

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language. By refunctionalization, the call stack of the analyzed language
is blended into the call stack of the defining language. This provides another perspective to
explain why \citeauthor{darais2017abstracting}'s abstract definitional interpreters
is able to extend the the pushdown control-flow property from its defining language.

@@

As mentioned in Section~\ref{unabs}, maintaining the perfect precision provided by an unbounded-stack machine is essential when performing the proposed transformations. Pushdown control flow analysis (PDCFA) is a mechanism which maintains this precision through the use of a specialized state graph representing all possible stacks contained within the unbounded-stack machine. These state graphs are known as \emph{Dyke state graphs}~\cite{gilray2016pushdown}.

>>>>

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, reynolds1972definitional}.
Nowadays, even an undergraduate student in computer science is able to build toy languages
through interpreters.
But building a sound abstract interpreter remained an esoteric and difficult task until very recently.

\citeauthor{van2012systematic} proposed the Abstracting Abstract Machines (AAM) methodology which provides a recipe
for constructing sound abstract interpreters for higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

Broadly speaking, abstract abstract machines and abstract definitional
interpreters are different forms of abstract interpreters.
They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interpreters, respectively.
An interesting question, and the subject of this paper, is how we can interderive these abstract semantics artifacts
from the respective other one.

In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters have been intensively studied by Danvy and his
collaborators
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The concrete abstract machines implement structural operational semantics in continuation-passing style,
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines,
and by defunctionalizing the higher-order functions, one may obtain abstract machines in the
reverse direction.

In the domain of abstract semantics artifacts, by contrast,
the relations between small-step abstract abstract machines and big-step abstract
definitional interpreters, as well as the question of deriving one from the other, are not
well studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we provide a constructive answer to the question 
of interderiving abstract semantics artifacts and relate AAM and ADI
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.

In addition, the abstract abstract machines with unbounded stack naturally correspond
to abstract definitional interpreters. We show that after refunctionalizing an AAM with unbounded
stack, and with the proper caching algorithm, the pushdown control-flow analysis can be
obtained.

@@

\subsection{Contributions}

@@

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language. By refunctionalization, the call stack of the analyzed language
is blended into the call stack of the defining language. This provides another perspective to
explain why \citeauthor{darais2017abstracting}'s abstract definitional interpreters
is able to inherit the pushdown control-flow property from its defining language.

@@

Pushdown control flow analysis (PDCFA) is another mechanism which maintains this precision 
through the use of a specialized state graph representing all possible stacks contained 
within the unbounded-stack machine. These state graphs are known as Dyke state graphs.
